---
title: "20260403_CRC_income_imputation"
output: html_document
date: "2026-04-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_chunk$set(echo = TRUE)

lapply(c('viridis', 'ggthemes', 'skimr', 'stringr', 'bigrquery', 'xgboost'),
       function(pkg_name) { if(! pkg_name %in% installed.packages()) { install.packages(pkg_name)} } )

library(tidyverse)  # Data wrangling packages.
library(bigrquery)  # Data extraction from Google BigQuery
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
library(xgboost)


```


```{r data import}
data_person <- fread("../data/20260325_072814_1268280261_data_person-000000000000.csv.gz")

anc_IDs <- fread("../data/clean_IDs.txt")
anc_IDs <- anc_IDs$V1
anc_list <- anc_IDs

data_person_anc <- data_person |>
  dplyr::filter(person_id %in% anc_list)

data_person_aou <- data_person |>
  dplyr::filter(!person_id %in% anc_list)

```


```{r}
survey_Occ0 <- fread("../data/20260325_073656_1885991969_data_surveyOccurrence-000000000000.csv.gz") |>
  dplyr::filter(person_id %in% anc_list)

files <- list.files(
  path = "../data",
  pattern = "surveyOccurrence-.*\\.csv\\.gz$",
  full.names = TRUE
)

# survey_anc <- rbindlist(lapply(files, function(f) {
#   dt <- fread(f)
#   dt[person_id %in% data_person_anc$person_id][, source_file := basename(f)]
# }))

survey_aou <- rbindlist(lapply(files, function(f) {
  dt <- fread(f)
  dt[person_id %in% data_person_aou$person_id][, source_file := basename(f)]
}))

#drop SDOH
survey_aou <- survey_aou |>
  dplyr::filter(T_DISP_survey_version %in% c("Lifestyle",
                                             "Overall Health",
                                             "The Basics"))

```


he cleanest XGBoost setup here is multiclass classification on the income bracket labels, not numeric regression on bracket midpoints. XGBoost tree models also handle missing predictor values by default, so your 5% missing education field can stay in the feature set.

I would not call plain XGBoost the universally “best” solution, though. For predictive fill-in, it is a strong baseline. For downstream survey inference, I would benchmark it against a proper multiple-imputation workflow. mice explicitly supports ordered categorical variables and uses polr by default for ordered factors with more than two levels. More broadly, model-based imputation relies on assumptions about the missingness mechanism: under MAR you condition on observed variables, but if income nonresponse is still driven by the unobserved income itself after conditioning on observables, that is MNAR and deserves sensitivity analysis rather than blind trust in one model.


Some Imputation Code
```{r}


# Keep this in the exact questionnaire order.
# If your real instrument has a "150k 200k" bucket, add it here in order.
income_levels <- c(
  "less 10k", "10k 25k", "25k 35k", "35k 50k",
  "50k 75k", "75k 100k", "100k 150k", "150k 200k", "more 200k"
)

coerce_xgb_types <- function(df) {
  out <- df
  for (nm in names(out)) {
    x <- out[[nm]]
    if (is.character(x)) {
      out[[nm]] <- factor(x)
    } else if (is.logical(x)) {
      out[[nm]] <- as.integer(x)
    } else if (inherits(x, "Date") || inherits(x, "POSIXt")) {
      out[[nm]] <- as.numeric(x)
    }
  }
  out
}

drop_constant_or_all_missing <- function(df, max_missing = 0.995) {
  keep <- vapply(df, function(x) {
    miss <- mean(is.na(x))
    nunq <- length(unique(x[!is.na(x)]))
    !(miss >= max_missing || nunq <= 1L)
  }, logical(1))
  df[, keep, drop = FALSE]
}

align_to_template <- function(new_df, template_df) {
  out <- new_df

  missing_cols <- setdiff(names(template_df), names(out))
  for (nm in missing_cols) out[[nm]] <- NA

  out <- out[, names(template_df), drop = FALSE]

  for (nm in names(template_df)) {
    if (is.factor(template_df[[nm]])) {
      out[[nm]] <- factor(
        out[[nm]],
        levels = levels(template_df[[nm]]),
        ordered = is.ordered(template_df[[nm]])
      )
    } else {
      out[[nm]] <- suppressWarnings(as.numeric(out[[nm]]))
    }
  }

  out
}

stratified_split <- function(y, p_train = 0.70, p_valid = 0.15, seed = 42) {
  stopifnot(p_train > 0, p_valid > 0, (p_train + p_valid) < 1)
  set.seed(seed)

  train <- valid <- test <- integer()
  idx_by_class <- split(seq_along(y), y, drop = TRUE)

  for (idx in idx_by_class) {
    idx <- sample(idx)
    n <- length(idx)

    n_train <- floor(p_train * n)
    n_valid <- floor(p_valid * n)

    if (n >= 3L) {
      n_train <- max(1L, n_train)
      n_valid <- max(1L, n_valid)
    } else if (n == 2L) {
      n_train <- 1L
      n_valid <- 0L
    } else {
      n_train <- 1L
      n_valid <- 0L
    }

    n_test <- n - n_train - n_valid
    if (n >= 3L && n_test < 1L) {
      n_test <- 1L
      if (n_train >= n_valid + 1L) {
        n_train <- n_train - 1L
      } else {
        n_valid <- max(0L, n_valid - 1L)
      }
    }

    train <- c(train, idx[seq_len(n_train)])

    if (n_valid > 0L) {
      valid <- c(valid, idx[seq.int(n_train + 1L, n_train + n_valid)])
    }

    start_test <- n_train + n_valid + 1L
    if (start_test <= n) {
      test <- c(test, idx[start_test:n])
    }
  }

  list(train = sort(train), valid = sort(valid), test = sort(test))
}

as_prob_matrix <- function(pred, nclass) {
  if (is.matrix(pred)) return(pred)
  matrix(pred, ncol = nclass, byrow = TRUE)
}

multiclass_logloss <- function(prob, y_true_zero_based, eps = 1e-15) {
  p <- pmax(pmin(prob, 1 - eps), eps)
  idx <- cbind(seq_along(y_true_zero_based), y_true_zero_based + 1L)
  -mean(log(p[idx]))
}

fit_income_bracket_imputer <- function(
  pool_df,
  income_col,
  income_levels,
  exclude_cols = character(),
  top_k = 80,
  seed = 42,
  params = list(
    objective = "multi:softprob",
    eval_metric = "mlogloss",
    max_depth = 6,
    eta = 0.05,
    min_child_weight = 5,
    subsample = 0.8,
    colsample_bytree = 0.8,
    tree_method = "hist"
  )
) {
  dat <- coerce_xgb_types(pool_df)
  dat[[income_col]] <- factor(dat[[income_col]], levels = income_levels, ordered = TRUE)

  observed <- !is.na(dat[[income_col]])
  if (sum(observed) < 1000L) {
    stop("Too few observed income rows to do a reliable train/valid/test workflow.")
  }

  x_all <- dat[, setdiff(names(dat), c(income_col, exclude_cols)), drop = FALSE]
  x_all <- drop_constant_or_all_missing(x_all)

  x_obs <- x_all[observed, , drop = FALSE]
  y_obs <- dat[[income_col]][observed]

  split_idx <- stratified_split(y_obs, seed = seed)

  x_train <- x_obs[split_idx$train, , drop = FALSE]
  x_valid <- align_to_template(x_obs[split_idx$valid, , drop = FALSE], x_train)
  x_test  <- align_to_template(x_obs[split_idx$test,  , drop = FALSE], x_train)

  y_train <- y_obs[split_idx$train]
  y_valid <- y_obs[split_idx$valid]
  y_test  <- y_obs[split_idx$test]

  dtrain <- xgb.DMatrix(x_train, label = as.integer(y_train) - 1L)
  dvalid <- xgb.DMatrix(x_valid, label = as.integer(y_valid) - 1L)

  params$num_class <- length(income_levels)

  # Pass 1: all features -> gain-based feature selection
  full_model <- xgb.train(
    params = params,
    data = dtrain,
    nrounds = 2000,
    watchlist = list(train = dtrain, valid = dvalid),
    early_stopping_rounds = 50,
    verbose = 1
  )

  imp <- as.data.frame(xgb.importance(model = full_model))
  imp <- imp[order(imp$Gain, decreasing = TRUE), , drop = FALSE]

  selected_features <- head(imp$Features, top_k)
  if (length(selected_features) == 0L) {
    selected_features <- colnames(x_train)
  }

  x_train_sel <- x_train[, selected_features, drop = FALSE]
  x_valid_sel <- x_valid[, selected_features, drop = FALSE]
  x_test_sel  <- x_test[,  selected_features, drop = FALSE]

  dtrain_sel <- xgb.DMatrix(x_train_sel, label = as.integer(y_train) - 1L)
  dvalid_sel <- xgb.DMatrix(x_valid_sel, label = as.integer(y_valid) - 1L)
  dtest_sel  <- xgb.DMatrix(x_test_sel,  label = as.integer(y_test)  - 1L)

  # Pass 2: refit on selected features only
  selected_model <- xgb.train(
    params = params,
    data = dtrain_sel,
    nrounds = 2000,
    watchlist = list(train = dtrain_sel, valid = dvalid_sel),
    early_stopping_rounds = 50,
    verbose = 1
  )

  prob_test <- as_prob_matrix(
    predict(selected_model, dtest_sel),
    nclass = length(income_levels)
  )

  pred_test <- factor(
    income_levels[max.col(prob_test, ties.method = "first")],
    levels = income_levels,
    ordered = TRUE
  )

  metrics <- list(
    mlogloss = multiclass_logloss(prob_test, as.integer(y_test) - 1L),
    accuracy = mean(pred_test == y_test),
    adjacent_accuracy = mean(abs(as.integer(pred_test) - as.integer(y_test)) <= 1L),
    mean_abs_bucket_error = mean(abs(as.integer(pred_test) - as.integer(y_test)))
  )

  best_iter <- attr(selected_model, "best_iteration")
  if (is.null(best_iter)) best_iter <- 2000L

  # Final refit on all observed-income rows using the chosen feature set
  x_obs_sel <- x_obs[, selected_features, drop = FALSE]
  x_obs_sel <- align_to_template(x_obs_sel, x_train_sel)
  dall <- xgb.DMatrix(x_obs_sel, label = as.integer(y_obs) - 1L)

  final_model <- xgb.train(
    params = params,
    data = dall,
    nrounds = best_iter,
    verbose = 0
  )

  list(
    model = final_model,
    income_col = income_col,
    income_levels = income_levels,
    selected_features = selected_features,
    feature_importance = imp,
    metrics = metrics,
    confusion = table(truth = y_test, pred = pred_test),
    x_template = x_obs_sel
  )
}

impute_income_bracket <- function(new_df, fit, m = 1, stochastic = TRUE, seed = 42) {
  dat <- coerce_xgb_types(new_df)

  if (!fit$income_col %in% names(dat)) {
    dat[[fit$income_col]] <- NA
  }

  dat[[fit$income_col]] <- factor(
    dat[[fit$income_col]],
    levels = fit$income_levels,
    ordered = TRUE
  )

  missing_income <- is.na(dat[[fit$income_col]])
  if (!any(missing_income)) {
    return(if (m == 1L) dat else replicate(m, dat, simplify = FALSE))
  }

  x_new <- align_to_template(dat, fit$x_template)
  dmiss <- xgb.DMatrix(x_new[missing_income, , drop = FALSE])

  prob <- as_prob_matrix(
    predict(fit$model, dmiss),
    nclass = length(fit$income_levels)
  )

  hard_class <- factor(
    fit$income_levels[max.col(prob, ties.method = "first")],
    levels = fit$income_levels,
    ordered = TRUE
  )

  point_imputed <- dat
  point_imputed[[fit$income_col]][missing_income] <- hard_class

  if (m == 1L || !stochastic) {
    return(point_imputed)
  }

  set.seed(seed)
  out <- vector("list", m)

  for (i in seq_len(m)) {
    draw_idx <- apply(prob, 1, function(p) {
      p <- p / sum(p)
      sample.int(length(p), size = 1L, prob = p)
    })

    draw_class <- factor(
      fit$income_levels[draw_idx],
      levels = fit$income_levels,
      ordered = TRUE
    )

    tmp <- dat
    tmp[[fit$income_col]][missing_income] <- draw_class
    out[[i]] <- tmp
  }

  out
}
```


Clean up my data
```{r}
# person_id | question              | answer
# 1001          | age                        | 34
# 1001          | highest_education_level    | Bachelor's degree
# 1001          | annual_household_income    | 50k 75k
# 1002          | age                        | 51

survey_aou_clean <- survey_aou |>
  dplyr::select(person_id,
                question,
                answer)


first_non_missing <- function(x) {
  x <- trimws(as.character(x))
  x <- x[!is.na(x) & x != ""]
  if (length(x) == 0L) NA_character_ else x[1]
}

to_binary_int <- function(x) {
  x <- tolower(trimws(as.character(x)))
  case_when(
    x %in% c("1", "yes", "y", "true", "checked", "selected") ~ 1L,
    x %in% c("0", "no", "n", "false", "unchecked", "not selected") ~ 0L,
    TRUE ~ NA_integer_
  )
}

survey_long_to_wide <- function(
  long_df,
  id_col = "person_id",
  question_col = "question",
  answer_col = "answer",
  keep_cols = character(),   # e.g. weights/strata/psu if repeated on each row
  missing_tokens = c(
    "", "NA", "N/A", "Skipped", "Refused",
    "Don't know", "Prefer not to answer"
  ),
  numeric_questions = character(),
  binary_questions = character(),
  ordered_questions = list(),
  dedupe = c("error", "first_non_missing", "collapse")
) {
  dedupe <- match.arg(dedupe)

  df <- long_df %>%
    mutate(
      .id = .data[[id_col]],
      .question = as.character(.data[[question_col]]),
      .answer = trimws(as.character(.data[[answer_col]])),
      .answer = ifelse(.answer %in% missing_tokens, NA_character_, .answer)
    ) %>%
    select(all_of(keep_cols), .id, .question, .answer)

  duplicates <- df %>%
    count(.id, .question, name = "n") %>%
    filter(n > 1)

  if (nrow(duplicates) > 0 && dedupe == "error") {
    stop(
      "Found duplicate respondent/question pairs. Inspect `duplicates` ",
      "or rerun with dedupe = 'first_non_missing' or 'collapse'."
    )
  }

  values_fun <- switch(
    dedupe,
    error = NULL,
    first_non_missing = list(.answer = first_non_missing),
    collapse = list(.answer = ~ paste(unique(stats::na.omit(.x)), collapse = " | "))
  )

  wide <- df %>%
    pivot_wider(
      id_cols = c(all_of(keep_cols), .id),
      names_from = .question,
      values_from = .answer,
      values_fill = NA_character_,
      values_fn = values_fun
    ) %>%
    rename(person_id = .id)

  if (length(numeric_questions) > 0) {
    wide <- wide %>%
      mutate(across(any_of(numeric_questions), ~ suppressWarnings(as.numeric(.x))))
  }

  if (length(binary_questions) > 0) {
    wide <- wide %>%
      mutate(across(any_of(binary_questions), to_binary_int))
  }

  if (length(ordered_questions) > 0) {
    for (nm in names(ordered_questions)) {
      if (nm %in% names(wide)) {
        wide[[nm]] <- factor(
          wide[[nm]],
          levels = ordered_questions[[nm]],
          ordered = TRUE
        )
      }
    }
  }

  list(
    wide = wide,
    duplicates = duplicates
  )
}

income_levels <- c(
  "less 10k", "10k 25k", "25k 35k", "35k 50k",
  "50k 75k", "75k 100k", "100k 150k", "more 200k"
)

edu_levels <- c("Never Attended",
                "One Through Four",
                "Five Through Eight",
                "Nine Through Eleven",
                "Twelve Or GED",
                "College One to Three",
                "College Graduate",
                "Advanced Degree")

pool_prep <- survey_long_to_wide(
  long_df = survey_aou_clean,
  id_col = "person_id",
  question_col = "question",
  answer_col = "answer",
  keep_cols = c("survey_weight", "strata", "psu"),  # only if these repeat per row
  numeric_questions = c("age", "household_size", "num_children"),
  binary_questions = c("employed", "married", "has_children"),
  ordered_questions = list(
    highest_education_level = edu_levels,
    annual_household_income = income_levels
  ),
  dedupe = "error"
)

wide_pool_df <- pool_prep$wide

# Inspect this if the function stops:
pool_prep$duplicates

fit <- fit_income_bracket_imputer(
  pool_df = wide_pool_df,
  income_col = "annual_household_income",
  income_levels = income_levels,
  exclude_cols = c("person_id")
)

mi_5k <- impute_income_bracket(
  new_df = wide_5k_df,
  fit = fit,
  m = 5,
  stochastic = TRUE,
  seed = 42
)

multi_select_wide <- multi_select_long %>%
  mutate(selected = 1L) %>%
  pivot_wider(
    id_cols = person_id,
    names_from = c(question, option_code),
    values_from = selected,
    values_fill = 0L
  )

wide_pool_df <- wide_pool_df %>%
  left_join(multi_select_wide, "person_id")
```



run the imputation
```{r}

# If your real instrument has a "150k 200k" bucket, add it here in order.
income_levels <- c(
  "less 10k", "10k 25k", "25k 35k", "35k 50k",
  "50k 75k", "75k 100k", "100k 150k", "150k 200k", "more 200k"
)


fit <- fit_income_bracket_imputer(
  pool_df = study_pool_df,
  income_col = "annual_household_income",
  income_levels = income_levels,
  exclude_cols = c(
    "person_id",
    "free_text_comment"
    # add anything derived from income itself
  ),
  top_k = 80,
  seed = 42
)

fit$metrics
head(fit$feature_importance, 20)
fit$confusion

# Deterministic single imputation
imputed_5k <- impute_income_bracket(
  new_df = analytic_5k_df,
  fit = fit,
  m = 1
)

# 5 stochastic completed datasets
mi_5k <- impute_income_bracket(
  new_df = analytic_5k_df,
  fit = fit,
  m = 5,
  stochastic = TRUE,
  seed = 42
)
```

