---
title: "20260325_CRC_ancill_demo.Rmd"
output: html_document
date: "2026-03-25"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)

lapply(c('viridis', 'ggthemes', 'skimr', 'stringr', 'bigrquery'),
       function(pkg_name) { if(! pkg_name %in% installed.packages()) { install.packages(pkg_name)} } )

library(stringr)
library(viridis)    # A nice color scheme for plots.
library(ggthemes)   # Common themes to change the look and feel of plots.
library(lubridate)  # Date library from the tidyverse.
library(tidyverse)  # Data wrangling packages.
library(bigrquery)  # Data extraction from Google BigQuery

library(data.table)
##library(zipcodeR) ### trying a different way

library(dplyr)
library(ggplot2)
library(purrr)
library(tidyr)
library(stringr)

```


```{r data import}
data_person <- fread("../data/20260325_072814_1268280261_data_person-000000000000.csv.gz")

waist <- fread("../data/20260325_073656_102899057_data_waistCircumference-000000000000.csv.gz")
hip <- fread("../data/20260325_073656_161696072_data_hipCircumference-000000000000.csv.gz")
height <- fread("../data/20260325_073656_454031082_data_height-000000000000.csv.gz")
weight <- fread("../data/20260325_073656_944882696_data_weight-000000000000.csv.gz")
bp <- fread("../data/20260325_073656_1405775660_data_bloodPressure-000000000000.csv.gz")
bmi <- fread("../data/20260325_073656_1406089081_data_bmi-000000000000.csv.gz")
zip <- fread("../data/20260325_073656_2143096321_data_zipcodeSocioeconomic-000000000000.csv.gz")

#genetic data
anc_pred <- fread("../../vwb-aou-datasets-controlled/v8/wgs/short_read/snpindel/aux/ancestry/echo_v4_r2.ancestry_preds.tsv")
#rels <- fread("../../vwb-aou-datasets-controlled/v8/wgs/short_read/")

```

# Random Sample
### pull a random sample of people
### these all have EHR and genomics
### later we will add in the IDs of the ancillary study here
```{r random sample}
anc_N <- 5626

rand_samp <- sample(data_person$person_id, size = anc_N)

anc_IDs <- fread("../data/clean_IDs.txt")
anc_IDs <- anc_IDs$V1

anc_list <- anc_IDs

```

### get the basics on everyone
```{r thin data}
#basic demo
anc_data_person <- data_person |>
  dplyr::filter(person_id %in% anc_list) |>
  dplyr::select(-ethnicity, -gender, -race) |>
  dplyr::rename(gender = T_DISP_gender,
                ethnicity = T_DISP_ethnicity,
                race = T_DISP_race,
                sex = T_DISP_sex_at_birth,
                self_report = T_DISP_self_reported_category) |>
  dplyr::mutate(dob = as.Date(date_of_birth)) |>
  dplyr::select(person_id,
                dob,
                gender,
                race,
                sex,
                ethnicity,
                self_report)
```

### get the physical measurements
```{r thin and merge data}
# 1. Define your dataframes and the desired suffixes
# Using a named list makes automation much easier
raw_data_list <- list(
  bp = bp, waist = waist, hip = hip, bmi = bmi, 
  ht = height, wt = weight
)

# 1. Prep each file: Filter by ID, Normalize Dates, and Add Suffixes + Flags
prep_list <- purrr::imap(raw_data_list, function(df, name) {
  df |> 
    dplyr::filter(person_id %in% anc_list) |> 
    dplyr::mutate(
      person_id = as.character(person_id),
      date = as.Date(date),
      # Add a hidden flag to prove this person had a record in this file on this date
      !!paste0("present.", name) := TRUE 
    ) |> 
    dplyr::rename_with(~paste0(., ".", name), -c(person_id, date, starts_with("present.")))
})

# 2. Join into one wide table
master_joined <- prep_list |> 
  purrr::reduce(dplyr::full_join, by = c("person_id", "date"))

# 3. Filter and Slice based on "Presence" rather than "Values"
final_df <- master_joined |>
  # Count actual non-NA measurements for the "richest data" tie-breaker
  # (We exclude the presence flags from this count)
  dplyr::mutate(
    data_points_count = rowSums(!is.na(dplyr::across(-c(person_id, date, starts_with("present."))))),
    # Ensure the "present" flags are treated as FALSE if they are NA after the join
    dplyr::across(starts_with("present."), ~dplyr::coalesce(., FALSE))
  ) |>
  # Logic: Keep if the record existed in Cat 2 AND Cat 1, even if values are NA
  dplyr::filter(
    (present.ht | present.wt) & 
    (present.bp | present.waist | present.hip | present.bmi)
  ) |>
  # 4. Pick the row with the most actual data
  dplyr::group_by(person_id) |>
  dplyr::slice_max(data_points_count, n = 1, with_ties = FALSE) |>
  dplyr::ungroup()

# How many of your 6k actually had matching dates?
length(unique(final_df$person_id))

# See which columns are the most complete
colSums(!is.na(final_df))

final_df |>
  dplyr::select(
    BP     = systolic.bp,
    Waist  = value_numeric.waist,
    Hip    = value_numeric.hip,
    BMI    = value_numeric.bmi,
    Height = value_numeric.ht,
    Weight = value_numeric.wt
  ) |>
  # Create a logical table (TRUE if data exists) and sum the TRUES
  is.na() |>
  (\(x) !x)() |> 
  colSums()
```

# Who is missing?
```{r}
# Find who is still missing from the final result
still_missing <- setdiff(anc_list, final_df$person_id)

# Check their presence flags in the master join
master_joined |> 
  filter(person_id %in% still_missing) |> 
  select(person_id, date, starts_with("present."))


missing_anc <- anc_list[!anc_list %in% final_df$person_id]

pick <- 174
#pick <- 224

missing_anc[pick]

data_person |> dplyr::filter(person_id == missing_anc[pick])
weight |> dplyr::filter(person_id == missing_anc[pick])
height |> dplyr::filter(person_id == missing_anc[pick])
hip |> dplyr::filter(person_id == missing_anc[pick])
bmi |> dplyr::filter(person_id == missing_anc[pick])
bp |> dplyr::filter(person_id == missing_anc[pick])

```

#final merge
```{r}
final_df_merge <- final_df |>
  dplyr::mutate(person_id = as.integer(person_id)) |>
  dplyr::left_join(zip, by = "person_id")

anc_data <- anc_data_person |>
  dplyr::left_join(final_df_merge, by = "person_id")
```


# Geographic location (Zip3) with State and Census Region
```{r}
# Your raw data
my_data <- data.frame(zip_raw = c("277**", "902**", "606**", "021**"))

# Remove asterisks to get a clean 3-digit string
my_data <- my_data %>%
  mutate(zip3 = str_remove_all(zip_raw, fixed("*")))

## a simple mapping function
get_region <- function(z) {
  z <- as.numeric(z)
  case_when(
    z >= 010 & z <= 089 ~ "NE", # CT, MA, ME, NH, RI, VT, NJ
    z >= 100 & z <= 199 ~ "NE", # NY, PA, DE
    z >= 200 & z <= 299 ~ "S",  # DC, MD, VA, WV, NC, SC
    z >= 300 & z <= 399 ~ "S",  # GA, FL, AL, MS, TN
    z >= 400 & z <= 499 ~ "MW", # KY, OH, IN, MI
    z >= 500 & z <= 599 ~ "MW", # IA, WI, MN, SD, ND, MT
    z >= 600 & z <= 699 ~ "MW", # IL, MO, KS, NE
    z >= 700 & z <= 799 ~ "S",  # LA, AR, OK, TX
    z >= 800 & z <= 899 ~ "W",  # CO, WY, ID, UT, AZ, NM, NV
    z >= 900 & z <= 999 ~ "W",  # CA, HI, OR, WA, AK
    TRUE ~ "Unknown"
  )
}

my_data <- my_data %>%
  mutate(region = get_region(zip3))

print(my_data)
```

# Get Region and State for Anc List
```{r}
anc_data <- anc_data |>
    dplyr::rename(zip_raw = zip_code) |>
    dplyr::mutate(zip3 = str_remove_all(zip_raw, fixed("*"))) |> 
    dplyr::mutate(region = get_region(zip3))
```


# Some Stats
```{r}
# Standard fix for the sex column
anc_data |>
  dplyr::mutate(sex = dplyr::case_match(sex,
    "Male"   ~ "M",
    "Female" ~ "F",
    .default = "Other"
  )) |>
  # To get a summary, you usually want to group by the category first
  dplyr::group_by(sex) |> 
  dplyr::summarise(count = dplyr::n(),
                   percent = 100 * dplyr::n()/nrow(anc_data))

anc_data |>
  dplyr::mutate(self_report = dplyr::case_match(self_report,
    "American Indian or Alaska Native"   ~ "American Indian or Alaska Native",
    "Asian"   ~ "Asian",
    "Black or African American"   ~ "Black or African American",
    "Hispanic"   ~ "Hispanic",
    "More than one population"   ~ "More than one population",
    "White"   ~ "White",
    .default = "Other"
  )) |>
  # To get a summary, you usually want to group by the category first
  dplyr::group_by(self_report) |> 
  dplyr::summarise(count = dplyr::n(),
                   percent = 100 * dplyr::n()/nrow(anc_data)) |>
  dplyr::arrange(desc(percent))

# 1. Create the Summary Table
sex_summary <- anc_data |>
  dplyr::mutate(sex = dplyr::case_match(sex,
    "Male"   ~ "M",
    "Female" ~ "F",
    .default = "Other"
  )) |>
  filter(!is.na(sex), !is.na(region)) |> # Remove any NAs for a cleaner view
  group_by(region, sex) |>
  summarise(count = n(), .groups = 'drop') |>
  group_by(region) |>
  mutate(percentage = (count / sum(count)) * 100)

print(sex_summary)

# 2. Visualize the Breakdown
# A stacked bar chart showing the percentage of T2D status per region
ggplot(sex_summary, aes(x = region, y = percentage, fill = as.factor(sex))) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_brewer(palette = "Set2", name = "Sex") +
  labs(
    title = "Breakdown of Sex by Census Region",
    x = "Census Region",
    y = "Percentage (%)",
    caption = "Data categorized by Zip3 mapping"
  ) +
  theme_minimal()
```

# Region Breakdown
```{r}

# --- STEP 3: Breakdown of Demographics by Region ---

# A. Age Distribution by Region
# (Assuming 'age' is a numeric column)
age_plot <- ggplot(anc_data, aes(x = region, y = age, fill = region)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Age Distribution by Census Region (Non-T2D)", y = "Age") +
  theme_minimal()

# B. Sex at Birth Breakdown
sex_breakdown <- anc_data %>%
  count(region, sex) %>%
  group_by(region) %>%
  mutate(pct = n / sum(n) * 100)

# C. Race Breakdown
race_breakdown <- anc_data %>%
  count(region, race) %>%
  group_by(region) %>%
  mutate(pct = n / sum(n) * 100)

# Print the tables
print("Sex at Birth by Region:")
print(sex_breakdown)

print("Race by Region:")
print(race_breakdown)

# Visualizing Race by Region
ggplot(race_breakdown, aes(x = region, y = pct, fill = race)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Race Breakdown by Census Region", y = "Percentage of Regional Sample") +
  theme_minimal()

# Create a combined Violin + Boxplot
ggplot(anc_data, aes(x = region, y = age_at_occurrence.bp, fill = region)) +
  geom_violin(alpha = 0.3) +
  geom_boxplot(width = 0.2, color = "black", outlier.shape = NA) +
  # Adding a jitter to see the actual data points (useful if sample size is small)
  geom_jitter(width = 0.1, alpha = 0.1, size = 0.5) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "Age Distribution by Census Region",
    subtitle = "Non-T2D Patients Only",
    x = "Region",
    y = "Age (Years)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Statistical test for age differences across regions
age_anova <- aov(age_at_occurrence.bp ~ region, data = anc_data)
summary(age_anova)

# If the p-value is < 0.05, you can see which specific regions differ:
# TukeyHSD(age_anova)

# 1. Chi-square for Sex at Birth vs. Region
sex_table <- table(anc_data$region, anc_data$sex)
sex_chisq <- chisq.test(sex_table)

print("Chi-square Test: Sex at Birth by Region")
print(sex_chisq)

# 2. Chi-square for Race vs. Region
race_table <- table(anc_data$region, anc_data$race)
race_chisq <- chisq.test(race_table)

print("Chi-square Test: Race by Region")
print(race_chisq)
```




# PCA Data from entire AoU
```{r}
table(anc_pred$ancestry_pred)

anc_pred_clean <- anc_pred |>
  dplyr::mutate(
    # 1. Remove the brackets [ ] so we just have "0.101, 0.138, ..."
    probabilities = str_remove_all(probabilities, "\\[|\\]")
  ) |>
  # 2. Split into multiple columns
  # names_sep = "_" will name them column_to_split_1, column_to_split_2, etc.
  separate_wider_delim(
    cols = probabilities,
    delim = ", ",
    names_sep = "_",
    too_few = "align_start"
  ) |>
  # 3. Convert all those new columns from character to numeric
  dplyr::mutate(across(starts_with("probabilities_"), as.numeric)) |>
  dplyr::rename(afr = probabilities_1,
         amr = probabilities_2,
         eas = probabilities_3,
         eur = probabilities_4,
         mid = probabilities_5,
         sas = probabilities_6) |>
  dplyr::rename(pca = pca_features) |>
  dplyr::mutate(
    # 1. Remove the brackets [ ] so we just have "0.101, 0.138, ..."
    pca = str_remove_all(pca, "\\[|\\]")
  ) |>
  separate_wider_delim(
    cols = pca,
    delim = ", ",
    names_sep = "_",
    too_few = "align_start"
  ) |>
  dplyr::mutate(across(starts_with("pca_"), as.numeric))

ancestry_anc <- anc_pred_clean |>
  dplyr::rename(person_id = research_id) |>
  dplyr::filter(person_id %in% anc_list)

### ggplot some 
ggplot(data = ancestry_anc,
       aes(x = pca_1,
           y = pca_2,
           col = ancestry_pred)) +
  geom_point() +
  theme_minimal()


ancestry_anc |>
  dplyr::left_join(anc_data, by = "person_id") |>
  dplyr::select(ancestry_pred, self_report) |>
  dplyr::group_by(ancestry_pred, self_report) |>
  dplyr::summarize(n = n()) 

ancestry_anc |>
  left_join(anc_data, by = "person_id") |>
  select(ancestry_pred, self_report) |>
  group_by(ancestry_pred, self_report) |>
  summarize(n = n(), .groups = "drop") |>
  pivot_wider(names_from = self_report,
              values_from = n,
              values_fill = 0)

ggplot(data = ancestry_anc |>
  left_join(anc_data, by = "person_id"),
       aes(x = pca_1,
           y = pca_2,
           col = self_report)) +
  geom_point() +
  theme_minimal()
```





# Get Genotype data from AoU

```{r}

# dim(rels)
# summary(rels$kin)
# rel_list
# 
# sum(anc_list %in% rel_list)
# sum(anc_list %in% rels$i.s)
# sum(anc_list %in% rels$j.s)
# 
# rels |>
#   dplyr::filter(i.s %in% rel_list$sample_id)
# 
# rels |>
#   dplyr::filter(j.s %in% rel_list$sample_id)


```

# Subset the PLINK calls
```{r}

## this is where we can check for X/Y and confirm sex
#bim <- fread("../../vwb-aou-datasets-controlled/v8/microarray/plink/arrays.bim")
#bed <- fread("../../vwb-aou-datasets-controlled/v8/microarray/plink/arrays.bed")
fam <- fread("../../vwb-aou-datasets-controlled/v8/microarray/plink/arrays.fam")

#head(bim)
head(fam)

# IDs you want to keep: one IID per line in ids_to_keep.txt
ids <- anc_list

# Read the PLINK fam file
#fam <- read.table("large_cohort.fam", header = FALSE)
colnames(fam)[1:2] <- c("FID", "IID")

# Match requested IDs to the source file
keep <- fam[fam$IID %in% ids, c("FID", "IID")]

# Check for missing IDs
missing_ids <- setdiff(ids, fam$IID)
length(missing_ids)
head(missing_ids)

# Write PLINK keep file
write.table(
  keep,
  file = "../data/plink/anc_list.txt",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE,
  sep = "\t"
)

```

# check the sex with plink
# run these commands in terminal mode

# 0) get to where you're going and make a local output directory on the VM
cd /home/rstudio/workspace/AncillaryDemographics/rmd
mkdir -p /tmp/plink_out

# 1) subset chrX to your keep-list into /tmp (about 5-10 mins)
# chrX: subset + split PAR + write to /tmp
plink2 \
  --threads 8 \
  --memory 60000 \
  --pfile ../../vwb-aou-datasets-controlled/v8/wgs/short_read/snpindel/acaf_threshold/pgen/acaf_threshold.chrX \
  --keep ../data/plink/anc_list.txt \
  --split-par hg38 \
  --make-pgen \
  --out /tmp/plink_out/anc_list_chrX

# 3) run chrX sex check
# this produces .sexcheck with IID, PEDSEX, SNPSEX, STATUS, F
plink2 \
  --threads 8 \
  --memory 60000 \
  --pfile /tmp/plink_out/anc_list_chrX \
  --check-sex max-female-xf=0.2 min-male-xf=0.8 \
  --out /tmp/plink_out/anc_list_chrX_sexcheck

# 4) subset chrY to your keep-list into /tmp
# chrX: subset + split PAR + write to /tmp
plink2 \
  --threads 8 \
  --memory 60000 \
  --pfile ../../vwb-aou-datasets-controlled/v8/wgs/short_read/snpindel/acaf_threshold/pgen/acaf_threshold.chrY \
  --keep ../data/plink/anc_list.txt \
  --make-pgen \
  --out /tmp/plink_out/anc_list_chrY


# 5) run chrY sample missingness
# this produces .smiss with IID, MISSING_CT, OBS_CT, F_MISS
plink2 \
  --threads 8 \
  --memory 60000 \
  --pfile /tmp/plink_out/anc_list_chrY \
  --missing sample-only \
  --y-nosex-missing-stats \
  --out /tmp/plink_out/anc_list_chrY_missing

# 6) inspect the two headers
head -n 5 /tmp/plink_out/anc_list_chrX_sexcheck.sexcheck
head -n 5 /tmp/plink_out/anc_list_chrY_missing.smiss

# 7) join the two files by IID into one tab-delimited table
awk 'BEGIN{FS=OFS="\t"}
FNR==1 && NR==1 {next}
FNR==1 {next}
NR==FNR {x[$1]=$2 OFS $3 OFS $4 OFS $5; next}
($1 in x) {print $1, x[$1], $2, $3, $4}
' /tmp/plink_out/anc_list_chrX_sexcheck.sexcheck /tmp/plink_out/anc_list_chrY_missing.smiss \
| awk 'BEGIN{FS=OFS="\t"; print "IID","PEDSEX","SNPSEX","STATUS","X_F","Y_MISSING_CT","Y_OBS_CT","Y_F_MISS"} {print}' \
> /tmp/plink_out/anc_list_chrXY_joined.tsv


# 7) preview the merged result
head -n 20 /tmp/plink_out/anc_list_chrXY_joined.tsv

cp /tmp/plink_out/anc_list_chrXY_joined.tsv ../data/plink/


```{r}
plink_sex <- fread("../data/plink/anc_list_chrXY_joined.tsv")

plink_sex <- plink_sex |>
  dplyr::rename(person_id = IID) |>
  dplyr::mutate(sex_pred_plink = ifelse(X_F > .85 & Y_F_MISS < .825, "M", "F"))

```


```{r}
plink_sex |>
  dplyr::filter(X_F > .8 & Y_F_MISS < .25)

plink_sex |>
  dplyr::filter(X_F < .5 & Y_F_MISS > .75)

ggplot(data = plink_sex,
       aes(x = X_F,
           y = Y_F_MISS)) +
  geom_point()


ggplot(data = plink_sex |>
        dplyr::left_join(anc_data, by = "person_id"),
       aes(x = X_F,
           y = Y_F_MISS,
           col = sex)) +
  geom_point()

ggplot(data = plink_sex |>
        dplyr::left_join(anc_data, by = "person_id"),
       aes(x = X_F,
           y = Y_F_MISS,
           col = gender)) +
  geom_hline(yintercept = .825) +
  geom_vline(xintercept = .85) +
  geom_point()

#high y missingness
plink_sex |>
  dplyr::left_join(anc_data, by = "person_id") |>
  dplyr::filter(X_F > .8 & Y_F_MISS > .25) |>
  dplyr::select(gender, sex) |>
  dplyr::mutate(g_s = paste0(gender,"-",sex)) 
```

# finally, combine ancestry, and plink sex info with anc_data
```{r}
ancillary_meta_data <- anc_data |>
  dplyr::left_join(ancestry_anc, by = "person_id") |>
  dplyr::left_join(plink_sex, by = "person_id")

saveRDS(ancillary_meta_data,
        file = "../data/ancillary_meta_data.RDS")

```


# bring in survey results (mostly for SDOH)
```{r}
survey_Occ0 <- fread("../data/20260325_073656_1885991969_data_surveyOccurrence-000000000000.csv.gz") |>
  dplyr::filter(person_id %in% anc_list)
survey_Occ1 <- fread("../data/20260325_073656_1885991969_data_surveyOccurrence-000000000001.csv.gz") |>
  dplyr::filter(person_id %in% anc_list)

files <- list.files(
  path = "../data",
  pattern = "surveyOccurrence-.*\\.csv\\.gz$",
  full.names = TRUE
)

survey_occ <- rbindlist(lapply(files, function(f) {
  dt <- fread(f)
  dt[person_id %in% anc_list][, source_file := basename(f)]
}))

saveRDS(survey_occ,
        file = "../data/ancillary_survey_data.RDS")

unique(survey_occ$T_DISP_question)

survey_occ |> 
  dplyr::filter(T_DISP_question == "Have you ever used an electronic nicotine product, even one or two times? (Electronic nicotine products include e-cigarettes, vape pens, hookah pens, personal vaporizers and mods, e-cigars, e-pipes, and e-hookahs.)") |>
  dplyr::group_by(T_DISP_answer) |>
  dplyr::summarise(n = n())

survey_occ |> 
  dplyr::filter(T_DISP_question == "Have you smoked at least 100 cigarettes in your entire life? (There are 20 cigarettes in a pack.)" ) |>
  dplyr::group_by(T_DISP_answer) |>
  dplyr::summarise(n = n())

survey_occ |> 
  dplyr::filter(T_DISP_question == "Do you now use electronic nicotine products…") |>
  dplyr::group_by(T_DISP_answer) |>
  dplyr::summarise(n = n())


survey_occ |> 
  dplyr::filter(T_DISP_question == "Do you now smoke cigarettes every day, some days, or not at all?") |>
  dplyr::group_by(T_DISP_answer) |>
  dplyr::summarise(n = n())


## cross reference smoking and e-cigs
# 1. Filter for just the two questions you want
# 2. Pivot so each person has one row with two columns
# 3. Pipe directly into the table function
cross_tab <- survey_occ |>
  filter(T_DISP_question %in% c(
    "Do you now use electronic nicotine products…",
    "Do you now smoke cigarettes every day, some days, or not at all?"
  )) |>
  pivot_wider(
    id_cols = person_id, # <--- Replace 'ID' with your actual respondent ID column name
    names_from = T_DISP_question, 
    values_from = T_DISP_answer
  ) |>
  select(-person_id) |> # Remove the ID so it doesn't clutter the table
  table()

# View the result
cross_tab

# 1. Filter and Recode
survey_clean <- survey_occ |>
  filter(T_DISP_question %in% c(
    "Do you now use electronic nicotine products…",
    "Do you now smoke cigarettes every day, some days, or not at all?"
  )) |>
  mutate(T_DISP_answer = case_match(
    T_DISP_answer,
    c("Dont Know", "Prefer Not To Answer", "Skip") ~ "Skip",
    c("Every Day", "Some Days")                   ~ "Some",
    "Not At All"                                  ~ "None",
    .default = T_DISP_answer  # Keeps other answers as-is if they don't match
  ))

# 2. Pivot and Table
cross_tab <- survey_clean |>
  pivot_wider(
    id_cols = person_id,  # <--- Change 'ID' to your actual respondent ID column
    names_from = T_DISP_question, 
    values_from = T_DISP_answer
  ) |>
  select(-person_id) |>
  table()

# 3. View the matrix
cross_tab



survey_occ |>
  dplyr::filter(T_DISP_question == "What is your annual household income from all sources?") |>
  dplyr::group_by(T_DISP_answer) |>
  dplyr::summarise(n = n()) |>
  dplyr::arrange(desc(n))

survey_occ |>
  dplyr::filter(T_DISP_question == "What is the highest grade or year of school you completed?") |>
  dplyr::group_by(T_DISP_answer) |>
  dplyr::summarise(n = n()) |>
  dplyr::arrange(desc(n))

```


```{r}
survey_income <- survey_occ |>
  dplyr::filter(T_DISP_question == "What is your annual household income from all sources?") |>
  dplyr::rename(annual_household_income = T_DISP_answer) |>
  dplyr::select(person_id, annual_household_income)

survey_edu <- survey_occ |>
  dplyr::filter(T_DISP_question == "What is the highest grade or year of school you completed?") |>
  dplyr::rename(highest_edu = T_DISP_answer) |>
  dplyr::select(person_id, highest_edu)

survey_sdoh <- survey_income |>
  dplyr::left_join(survey_edu, by = "person_id")

head(survey_sdoh)

ancillary_meta_data <- ancillary_meta_data |>
  dplyr::left_join(survey_sdoh, by = "person_id")

saveRDS(ancillary_meta_data,
        file = "../data/ancillary_meta_data.RDS")

```

```{r}

table(survey_sdoh$annual_household_income)

table(survey_sdoh$highest_edu)

midpoint_map <- c(
  "less 10k" = 5,
  "10k 25k" = 17.5,
  "25k 35k" = 30,
  "35k 50k" = 42.5,
  "50k 75k" = 62.5,
  "75k 100k" = 87.5,
  "100k 150k" = 125,
  "150k 200k" = 175,
  "more 200k" = 225
)

education_years_map <- c(
  "Never Attended" = 0,
  "One Through Four" = 2.5,
  "Five Through Eight" = 6.5,
  "Nine Through Eleven" = 10,
  "Twelve Or GED" = 12,
  "College One to Three" = 14,
  "College Graduate" = 16,
  "Advanced Degree" = 18
)

just_sex <- ancillary_meta_data |>
  dplyr::select(person_id, sex_pred_plink)

sdoh <- survey_sdoh |>
  dplyr::mutate(income_num = recode(annual_household_income,
                                    !!!midpoint_map,
                                    .default = NA_real_)) |>
  dplyr::mutate(education_years = recode(highest_edu,
                                         !!!education_years_map,
                                         .default = NA_real_)) |>
  dplyr::left_join(just_sex, by = "person_id")

ggplot(data = sdoh |> dplyr::filter(!is.na(income_num)) |> dplyr::filter(!is.na(education_years)),
       aes(x = education_years,
           y = income_num,
           col = sex_pred_plink)) +
  geom_jitter(alpha = .05,
              height = 5,
              width = .2) +
  geom_smooth(method = "loess", se = FALSE) +
  theme_minimal()

```


# Sandbox Exploration
### multiple weights/heights?

```{r}
weight |> 
  dplyr::filter(person_id %in% anc_list) |> 
  dplyr::add_count(person_id) |> 
  dplyr::filter(n > 1) |>
  dplyr::arrange(person_id)

# 1. Prepare the data
plot_data <- weight |>
  dplyr::mutate(date = as.Date(date)) |>
  dplyr::filter(person_id %in% anc_list) |>
  dplyr::group_by(person_id) |>
  dplyr::arrange(date, .by_group = TRUE) |>
  dplyr::mutate(
    # Calculate difference from the previous weight
    time_between = as.numeric(date - dplyr::lag(date), units = "days"),
    weight_diff = value_numeric - dplyr::lag(value_numeric)
  ) |>
  # Remove the first row for each person (it will have NA for differences)
  dplyr::filter(!is.na(time_between)) |>
  dplyr::ungroup() |>
  dplyr::mutate(weight_change = ifelse(weight_diff > 0, "gain",
                                       ifelse(weight_diff < 0, "loss", "none")))

# 2. Create the plot
ggplot(plot_data, aes(x = time_between,
                      y = abs(weight_diff),
                      col = weight_change)) +
  geom_point(alpha = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") + # Baseline for no change
  geom_smooth(method = "lm",
              formula = "y~x",
              color = "black", se = FALSE) + 
  labs(
    title = "Weight Change vs. Time Interval",
    x = "Days Between Measurements",
    y = "Change in Weight (kg)"
  ) +
  theme_minimal()


### same for height
height |> 
  dplyr::filter(person_id %in% anc_list) |> 
  dplyr::add_count(person_id) |> 
  dplyr::filter(n > 1) |>
  dplyr::arrange(person_id)

# 1. Prepare the data
plot_data <- height |>
  dplyr::mutate(date = as.Date(date)) |>
  dplyr::filter(person_id %in% anc_list) |>
  dplyr::group_by(person_id) |>
  dplyr::arrange(date, .by_group = TRUE) |>
  dplyr::mutate(
    # Calculate difference from the previous weight
    time_between = as.numeric(date - dplyr::lag(date), units = "days"),
    height_diff = value_numeric - dplyr::lag(value_numeric)
  ) |>
  # Remove the first row for each person (it will have NA for differences)
  dplyr::filter(!is.na(time_between)) |>
  dplyr::ungroup() |>
  dplyr::mutate(height_change = ifelse(height_diff > 0, "gain",
                                       ifelse(height_diff < 0, "loss", "none")))

# 2. Create the plot
ggplot(plot_data, aes(x = time_between,
                      y = abs(height_diff),
                      col = height_change)) +
  geom_point(alpha = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") + # Baseline for no change
  geom_smooth(method = "lm",
              formula = "y~x",
              color = "black", se = FALSE) + 
  labs(
    title = "Height Change vs. Time Interval",
    x = "Days Between Measurements",
    y = "Change in Height (cm)"
  ) +
  theme_minimal()
```